---
title: "iblir Package Tutorial"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{iblir Package Tutorial}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
# avoid check where vignette filename must equal the vignette title
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r, include = FALSE}
pkg = 'iblir' #so we can easily switch names if we decide to rename
```

## Summary

This document provides a brief introduction to the **Introductory
Biostatistics Labs in R (iblir)** package.

## Information for users

### Getting Started

The package is not yet on CRAN, so you need to install it from GitHub as
follows.

    install.packages('remotes')
    library('remotes')
    install_github('andreashandel/iblir')

After the package is installed, you can load it

    library('iblir')

To start the tutorials/labs, you type

    run_tutorial('LabN', package = 'iblir')

where N is the lab number, e.g. `Lab1`, `Lab2`.

A startup message also tells you this.

The lab should open in your browser, and you can go through it.

Once finished with a lab, close the tab in your browser. Then go back to
`RStudio` and stop the lab by pressing the `Stop` sign in the upper
right corner of the `Background Jobs` tabs. You can then go back to the
`Console` tab and use `run_tutorial()` again to start another lab.

If you use RStudio, there is an alternative to calling `run_tutorial()`.
Once **`r pkg`** is loaded, you should see its tutorials listed in the
`Tutorial` tab in the upper right window.

You can click the `Start Tutorial` button to start a lab. It will open
in the built-in browser. You'll likely want to open the lab in a full
browser window by clicking the `Show in new window` icon in the top left
corner of the tutorial. Note that this method uses RStudio's built-in
browser, which is not always quite as good as a real web browser. So if
things don't fully work, switch to the method of using `run_tutorial()`
and make sure it opens in your real browser.

### Lab Content

LAB 1: Introduction to the online lab format and Introduction to R and
"Data".

LAB 2: Focuses on interpreting and making histograms in R. 1.
Interpreting Histograms 2. Which variable types to use with histograms
3. Making histograms 4. Impact of bin selection and 4 methods to help
get the right size

LAB 3: Focuses on interpreting and making box plots in R 1. Interpreting
Box plots 2. Making Box plots 3. Comparing box plots

LAB 4: Introduces students to the steps of a basic data analysis 1. Data
Cleaning 2. Variable types in R 3. Making Tables 4. Making Figures with
numeric summaries

LAB 5: Life Tables 1. Life expectancy 2. Survival Curves 3. Hazard
Functions

LAB 6: Diagnostic and Screening Tests 1. Diagnostic and Screening tests
2. Sensitivity, Specificity, and Accuracy 3. ROC curves 4. AUC
calculations 5. Cut-off values

LAB 7: Discrete Distributions 1. The Binomial Distribution a.
Calculating Probability and Percentiles 2. The Poisson Distribution a.
Calculating Probability and Percentiles

LAB 8: Normal Distribution 1. Calculating Probability and Percentiles

LAB 9: Random Sampling and Central Limit Theorem 1. Review of
distributions a. Setting to show that they can be approximated by normal
distribution 2. Random sampling a. So that getting good estimates with
random samples does not require very large samples in many cases 3.
Central Limit Theorem a. Apply it to the data in the exercises and see
if resulting distributions are normal

LAB 10: One Sample T Methods 1. 1-sample t-tests a. Hypothesis testing
and Confidence Intervals 2. Making new variables in R a. Using a
variable in the data set to make a new one

LAB 11: Paired t Test 1. Paired t-tests in R a. Hypothesis tests and
Confidence Intervals 2. Visualizing paired data in R

LAB 12: Two Sample t-test 1. Two-sample t-test in R a. Hypothesis tests
and Confidence Intervals 2. Violin Plots in R

LAB 13: Hypothesis Tests and Confidence Intervals for Proportions 1. One
Proportion tests in R a. Hypothesis tests and Confidence Intervals 2.
Two Proportion tests in R a. Hypothesis tests and Confidence Intervals

## Information for teachers

There are several ways **`r pkg`** can be used to teach the labs. Each
is briefly described below.

### Local

This way of using **`r pkg`** is described above. Simply install and
load the package. The greeting message will tell you how to load each
lab. The Package website also provides this information.


### Posit (formerly RStudio) Cloud

_This assumes you know how to set up projects in Posit Cloud. This is not very hard._

With this approach, students run `R` and `RStudio` in the cloud through their browser, and don't have to install anything on their local computer.

For this approach, set up a new [PositCloud](https://posit.cloud) project and install the `iblir`
package (which will also install a ton of others). Then give students
the link to the workspace. They log in, load the package and can use it
as if they were running it locally.

Note that a free RStudio Cloud account might not be enough and you
might need to upgrade to a paid account.



### shinyapps.io

_This assumes you have a shinyapps.io account and now how to deploy apps. This is not too complicated._


Follow the [shinyapps.io user
guide](https://docs.posit.co/shinyapps.io/) to set up your
account and configure `rsconnect`.

Next, load **`r pkg`** on your local computer and run the `deploy_labs()` function, which will send each lab to your
Shiny server. By default, all labs are deployed, but you can specify
which ones by providing the function with a vector of lab numbers, e.g.
`deploy_labs(labvec = c(2,5))` will deploy only labs 2 and 5. The labs
will be on shinyapps.io at URLs like this one :
`https://YOURACCOUNTNAME.shinyapps.io/lab1/` etc.



### Your own Shiny server

_This assumes you have your own Shiny server up and running. This requires a good bit of technical expertise. One option to run your own Shiny server is to use DigitalOcean and e.g.
the [analogsea](https://pacha.dev/analogsea/) package._


Load **`r pkg`**, then run the `get_labs()` function. It will open a
file browser and point you to a zip file with all the labs. Copy that
zip file to your shiny server, ideally into its own folder (e.g. a
folder called `iblir`). Then extract. You should end up with a folder
for each lab, e.g. `/srv/shiny-server/iblir/Lab1`,
`/srv/shiny-server/iblir/Lab2`, and so on. Make sure the `iblir` package
is installed on your server. Also, make sure permissions for the folders
are set correctly.

Once all is set up, students should be able to go to the Server/Lab URL
to access each lab, e.g., by going to
`shinyserver.myplace.edu/iblir/Lab1` and so on.



## Information for developers (or teachers)

If you want to make changes to the labs, here is some brief information to get you started. 

**If you enhance or add to the labs, we would love to receive your contributions and make them part of `r pkg`! For more see `Contributing to the Package` below.**


### Edit the package

The best way is if you edit the whole package. 

First, to make sure you have all needed package dependencies, install **`r pkg`** from GitHub as described above (no need to load it).

Now, fork **`r pkg`** on GitHub, then clone it to your local machine. Open the package repository by opening `iblir.Rproj` with RStudio. 

All labs are within the `/inst/tutorials/` folder of the package repository. Each consists of a main R Markdown (`.Rmd`) file, and as needed a folder `images` for images and a folder `www` for other files. If you want to modify an existing lab, open the corresponding `.Rmd` file and edit it. If you want to add a new lab, start a new folder and follow the content and structure of the existing labs to create a new one. Keep the naming consistent (i.e., call it `LabN` with `N` whatever the next number is).

As you make changes, run `Install` from the `Build` tab in the upper right pane of RStudio to rebuild and re-install your updated package. Test it to make sure everything works. You can also run the `Check` command to make sure the package is fine (you can ignore notes). If you use some new `R` packages for your lab, you need to open the `DESCRIPTION` file in the main folder and add the new package to the `Imports` list.

Once everything is ready and works, take all the labs, put them in a zip file, and replace the existing `alllabs.zip` inside the `/inst/allabs/` folder with your zip file.

Finally, push your updated package to GitHub. 

Students/teachers can now use your version of the package in exactly the same ways as described above, with the only difference that any install command should now point to your fork of the package (e.g., install via `install_github('yourgithubaccount/iblir')`).

As a last step, **please contribute your enhancements to the packag! See below.**


### Modify labs but not the package

Again, we recommend using the approach just described.
But if for some reason you are not comfortable editing the package, here is another way to make adjustments.

Start by loading **`r pkg`** and running the `get_labs()` function. 

Unzip the folder, find the lab you want to change, and make any changes you like. You can also add a new lab as described in the prior section.
Test your changes to make sure everything works.

Once all works, you can then copy your new labs to your own shiny server as described above. If you want to deploy to
shinyapps.io, call the `deploy_labs()` function with an additional input specifying the folder in which all your new labs are stored, e.g.,
`deploy_labs(labfolder =`C:\tmp\mylabs')\`.

**This only works if you deploy the labs to a Shiny server. For local use or through Posit Cloud, you need to use the previous method.**


### More information

Inside the repository, go to the `/auxiliary/docsfordevelopers` folder.



## Contributing to the package

If you made changes/improvements to the package as described above, please send a pull request. We would love to get contributions from others.

If you plan on making large changes, it might be best to file an issue first and let us know. Then we can discuss and make sure your planned modifications can be
seamlessly integrated.

If you don't feel comfortable editing the package, but have some other feedback, e.g., bug reports, feature requests, please file a [GitHub
issue](https://github.com/andreashandel/iblir/issues). Any contribution is very welcome.



## Further information

See the [package website](https://andreashandel.github.io/iblir/) for
useful additional information.

## Contributors

The main content of this package was developed by [Brian
McKay](https://brianmckay.netlify.app/). The package is maintained by
[Brian McKay](https://brianmckay.netlify.app/) and [Andreas
Handel](https://www.andreashandel.com/). A full list of contributors and
a Bibtex entry for the citation [can be found
here](https://andreashandel.github.io/iblir/authors.html).

## References
